---
title: "Hands-on_Ex2"
editor: visual
---

## Global and Local Measures of Spatial Autocorrelation

We will learn how to compute Global and Local Measure of Spatial Autocorrelation (GLSA) by using **spdep** package.

## Getting Started

### The Analytical Question

Our task in this study, hence, is to apply appropriate spatial statistical methods to discover if development are even distributed geographically.

In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e. GDP per capita) of Hunan Provice, People Republic of China.

### The Study Area and Data

2 datasets will be used in this hands-on exercise:

1.  Hunan province administrative boundary layer at county level. This is a geospatial data set in ESRI shapefile format.
2.  Hunan_2012.csv: This csv file contains selected Hunan\'s local development indicators in 2012.

### Setting the Analytical Tools

The code chunk below is used to perform the following tasks:

1.  Creating a package list containing the necessary R packages,
2.  Checking if the R packages in the package list have been installed in R,
    -   If they have yet to be installed, RStudio will installed the missing packages,
3.  Launching the packages into R environment.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

## Getting the Data Into R Environment

We will learn how to bring a geospatial data and its associated attribute table into R environment.

### Import shapefile into r environment

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### Import csv file into r environment

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv",show_col_types = FALSE)
```

### Performing relational join

The code chunk below will be used to update the attribute table of *hunan*\'s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe.

```{r}
hunan <- left_join(hunan,hunan2012)
```

### Visualising Regional Development Indicator

We are going to prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 by using *qtm()* of **tmap** package.

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## Global Spatial Autocorrelation

We will learn how to compute global spatial autocorrelation statistics and to perform spatial complete randomness test for global spatial autocorrelation.
