---
title: "Hands-on_Ex2"
editor: visual
---

## Global and Local Measures of Spatial Autocorrelation

We will learn how to compute Global and Local Measure of Spatial Autocorrelation (GLSA) by using **spdep** package.

## Getting Started

### The Analytical Question

Our task in this study, hence, is to apply appropriate spatial statistical methods to discover if development are even distributed geographically.

In this case study, we are interested to examine the spatial pattern of a selected development indicator (i.e. GDP per capita) of Hunan Provice, People Republic of China.

### The Study Area and Data

2 datasets will be used in this hands-on exercise:

1.  Hunan province administrative boundary layer at county level. This is a geospatial data set in ESRI shapefile format.
2.  Hunan_2012.csv: This csv file contains selected Hunan's local development indicators in 2012.

### Setting the Analytical Tools

The code chunk below is used to perform the following tasks:

1.  Creating a package list containing the necessary R packages,
2.  Checking if the R packages in the package list have been installed in R,
    -   If they have yet to be installed, RStudio will installed the missing packages,
3.  Launching the packages into R environment.

```{r}
pacman::p_load(sf, spdep, tmap, tidyverse)
```

## Getting the Data Into R Environment

We will learn how to bring a geospatial data and its associated attribute table into R environment.

### Import shapefile into r environment

The code chunk below uses [*st_read()*](https://r-spatial.github.io/sf/reference/st_read.html) of **sf** package to import Hunan shapefile into R. The imported shapefile will be **simple features** Object of **sf**.

```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```

### Import csv file into r environment

Next, we will import *Hunan_2012.csv* into R by using *read_csv()* of **readr** package.

```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv",show_col_types = FALSE)
```

### Performing relational join

The code chunk below will be used to update the attribute table of *hunan*'s SpatialPolygonsDataFrame with the attribute fields of *hunan2012* dataframe.

```{r}
hunan <- left_join(hunan,hunan2012)
```

### Visualising Regional Development Indicator

We are going to prepare a basemap and a choropleth map showing the distribution of GDPPC 2012 by using *qtm()* of **tmap** package.

```{r}
equal <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal interval classification")

quantile <- tm_shape(hunan) +
  tm_fill("GDPPC",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5) +
  tm_layout(main.title = "Equal quantile classification")

tmap_arrange(equal, 
             quantile, 
             asp=1, 
             ncol=2)
```

## Global Spatial Autocorrelation

We will learn how to compute global spatial autocorrelation statistics and to perform spatial complete randomness test for global spatial autocorrelation.

### Computing Contiguity Spatial Weights

The spatial weights is used to define the neighbourhood relationships between the geographical units (i.e. county) in the study area.

In the code chunk below, [*poly2nb()*](https://r-spatial.github.io/spdep/reference/poly2nb.html) of **spdep** package is used to compute contiguity weight matrices for the study area. This function builds a neighbours list based on regions with contiguous boundaries.

The code chunk below is used to compute Queen contiguity weight matrix.

```{r}
wm_q <- poly2nb(hunan, 
                queen=TRUE)
summary(wm_q)
```

### Row-standardised weights matrix

We need to assign weights to each neighboring polygon. In our case, each neighboring polygon will be assigned equal weight (style="W").

```{r}
rswm_q <- nb2listw(wm_q, 
                   style="W", 
                   zero.policy = TRUE)
rswm_q
```

### Global Spatial Autocorrelation: Moran's I

We will learn how to perform Moran's I statistics testing by using [*moran.test()*](https://r-spatial.github.io/spdep/reference/moran.test.html) of **spdep**.

### Maron's I test

The code chunk below performs Moran's I statistical testing using [*moran.test()*](https://r-spatial.github.io/spdep/reference/moran.test.html) of **spdep**.

```{r}
moran.test(hunan$GDPPC, 
           listw = rswm_q, 
           zero.policy = TRUE, 
           na.action=na.omit)
```

#### Computing Monte Carlo Moran's I

The code chunk below performs permutation test for Moran's I statistic by using [*moran.mc()*](https://r-spatial.github.io/spdep/reference/moran.mc.html) of **spdep**. A total of 1000 simulation will be performed.

```{r}
set.seed(1234)
bperm= moran.mc(hunan2012$GDPPC, 
                listw=rswm_q, 
                nsim=999, 
                zero.policy = TRUE, 
                na.action=na.omit)
bperm
```

#### Visualising Monte Carlo Moran's I

We will plot the distribution of the statistical values as a histogram by using the code chunk below.

In the code chunk below [*hist()*](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/hist) and [*abline()*](https://www.rdocumentation.org/packages/graphics/versions/3.6.2/topics/abline) of R Graphics are used.

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999]) 
```

```{r}
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, 
     freq=TRUE, 
     breaks=20, 
     xlab="Simulated Moran's I")
abline(v=0, 
       col="red") 
```

### Global Spatial Autocorrelation: Geary's

We will learn how to perform Geary's c statistics testing by using appropriate functions of **spdep** package.

The code chunk below performs Geary's C test for spatial autocorrelation by using [*geary.test()*](https://r-spatial.github.io/spdep/reference/geary.test.html) of **spdep**.

```{r}
geary.test(hunan$GDPPC, listw=rswm_q)
```

#### Computing Monte Carlo Geary's C

The code chunk below performs permutation test for Geary's C statistic by using [*geary.mc()*](https://r-spatial.github.io/spdep/reference/geary.mc.html) of **spdep**.

```{r}
set.seed(1234)
bperm=geary.mc(hunan$GDPPC, 
               listw=rswm_q, 
               nsim=999)
bperm
```

#### Visualising the Monte Carlo Geary's C

We will plot a histogram to reveal the distribution of the simulated values by using the code chunk below.

```{r}
mean(bperm$res[1:999])
```

```{r}
var(bperm$res[1:999])
```

```{r}
summary(bperm$res[1:999])
```

```{r}
hist(bperm$res, freq=TRUE, breaks=20, xlab="Simulated Geary c")
abline(v=1, col="red") 
```

## Spatial Correlogram

They show how correlated are pairs of spatial observations when you increase the distance (lag) between them - they are plots of some index of autocorrelation (Moran's I or Geary's c) against distance.

They are very useful as an exploratory and descriptive tool. For this purpose they actually provide richer information than variograms.

### Compute Moran's I correlogram

In the code chunk below, [*sp.correlogram()*](https://r-spatial.github.io/spdep/reference/sp.correlogram.html) of **spdep** package is used to compute a 6-lag spatial correlogram of GDPPC. The global spatial autocorrelation used in Moran's I. The **plot()** of base Graph is then used to plot the output.

```{r}
MI_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="I", 
                          style="W")
plot(MI_corr)

print(MI_corr)
```

### Compute Geary's C correlogram and plot

In the code chunk below, *sp.correlogram()* of **spdep** package is used to compute a 6-lag spatial correlogram of GDPPC. The global spatial autocorrelation used in Geary's C. The **plot()** of base Graph is then used to plot the output.

```{r}
GC_corr <- sp.correlogram(wm_q, 
                          hunan$GDPPC, 
                          order=6, 
                          method="C", 
                          style="W")
plot(GC_corr)

print(GC_corr)
```

## Cluster and Outlier Analysis

We will learn how to apply appropriate Local Indicators for Spatial Association (LISA), especially local Moran'I to detect cluster and/or outlier from GDP per capita 2012 of Hunan Province, PRC.

### Computing local Moran's I

To compute local Moran's I, the [*localmoran()*](https://r-spatial.github.io/spdep/reference/localmoran.html) function of **spdep** will be used. It computes *Ii* values, given a set of *zi* values and a listw object providing neighbour weighting information for the polygon associated with the zi values.

The code chunks below are used to compute local Moran's I of *GDPPC2012* at the county level.

```{r}
fips <- order(hunan$County)
localMI <- localmoran(hunan$GDPPC, rswm_q)
head(localMI)
```

The code chunk below list the content of the local Moran matrix derived by using [*printCoefmat()*](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/printCoefmat).

```{r}
printCoefmat(data.frame(localMI[fips,], row.names=hunan$County[fips]), check.names=FALSE)
```

#### Mapping the local Moran's I

The code chunks below can be used to perform the task. The out SpatialPolygonDataFrame is called *hunan.localMI*.

```{r}
hunan.localMI <- cbind(hunan,localMI) %>%
  rename(Pr.Ii = Pr.z....E.Ii..)
```

#### Mapping local Moran's I values

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty",
          palette = "RdBu",
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)
```

#### Mapping local Moran's I p-values

The code chunks below produce a choropleth map of Moran's I p-values by using functions of **tmap** package.

```{r}
tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)
```

#### Mapping both local Moran's I values and p-values

The code chunk below will be used to create such visualisation.

```{r}
localMI.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Ii", 
          style = "pretty", 
          title = "local moran statistics") +
  tm_borders(alpha = 0.5)

pvalue.map <- tm_shape(hunan.localMI) +
  tm_fill(col = "Pr.Ii", 
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette="-Blues", 
          title = "local Moran's I p-values") +
  tm_borders(alpha = 0.5)

tmap_arrange(localMI.map, pvalue.map, asp=1, ncol=2)
```

## Creating a LISA Cluster Map

The first step before we can generate the LISA cluster map is to plot the Moran scatterplot.

### Plotting Moran scatterplot

The code chunk below plots the Moran scatterplot of GDPPC 2012 by using [*moran.plot()*](https://r-spatial.github.io/spdep/reference/moran.plot.html) of **spdep**.

```{r}
nci <- moran.plot(hunan$GDPPC, rswm_q,
                  labels=as.character(hunan$County), 
                  xlab="GDPPC 2012", 
                  ylab="Spatially Lag GDPPC 2012")
```

### Plotting Moran scatterplot with standardised variable

We will use [*scale()*](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/scale) to centers and scales the variable. Here centering is done by subtracting the mean (omitting NAs) the corresponding columns, and scaling is done by dividing the (centered) variable by their standard deviations.

```{r}
hunan$Z.GDPPC <- scale(hunan$GDPPC) %>% as.vector 
```

The [*as.vector()*](https://www.rdocumentation.org/packages/pbdDMAT/versions/0.5-1/topics/as.vector) added to the end is to make sure that the data type we get out of this is a vector, that map neatly into out dataframe.

Now, we are ready to plot the Moran scatterplot again by using the code chunk below.

```{r}
nci2 <- moran.plot(hunan$Z.GDPPC, rswm_q,
                   labels=as.character(hunan$County),
                   xlab="z-GDPPC 2012", 
                   ylab="Spatially Lag z-GDPPC 2012")
```

### Preparing LISA map classes

The code chunks below show the steps to prepare a LISA cluster map.

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
```

We centers the variable of interest around its mean.

```{r}
DV <- hunan$GDPPC - mean(hunan$GDPPC)     
```

Next, we center the local Moran's around the mean.

```{r}
C_mI <- localMI[,1] - mean(localMI[,1])    
```

We will set a statistical significance level for the local Moran.

```{r}
signif <- 0.05       
```

These four command lines define the high-high, low-low, low-high and high-low categories.

```{r}
quadrant[DV >0 & C_mI>0] <- 4      
quadrant[DV <0 & C_mI<0] <- 2      
quadrant[DV <0 & C_mI>0] <- 1
quadrant[DV >0 & C_mI<0] <- 3
```

Lastly, we place non-significant Moran in the category 0.

```{r}
quadrant[localMI[,5]>signif] <- 0
```

We can combined all the steps into one single code chunk as shown below:

```{r}
quadrant <- vector(mode="numeric",length=nrow(localMI))
DV <- hunan$GDPPC - mean(hunan$GDPPC)     
C_mI <- localMI[,1] - mean(localMI[,1])    
signif <- 0.05       
quadrant[DV >0 & C_mI>0] <- 4      
quadrant[DV <0 & C_mI<0] <- 2      
quadrant[DV <0 & C_mI>0] <- 1
quadrant[DV >0 & C_mI<0] <- 3
quadrant[localMI[,5]>signif] <- 0
```

### Plotting LISA map

We can build the LISA map by using the code chunks below.

```{r}
hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)
```

For effective interpretation, it is better to plot both the local Moran's I values map and its corresponding p-values map next to each other.

The code chunk below will be used to create such visualisation.

```{r}
gdppc <- qtm(hunan, "GDPPC")

hunan.localMI$quadrant <- quadrant
colors <- c("#ffffff", "#2c7bb6", "#abd9e9", "#fdae61", "#d7191c")
clusters <- c("insignificant", "low-low", "low-high", "high-low", "high-high")

LISAmap <- tm_shape(hunan.localMI) +
  tm_fill(col = "quadrant", 
          style = "cat", 
          palette = colors[c(sort(unique(quadrant)))+1], 
          labels = clusters[c(sort(unique(quadrant)))+1],
          popup.vars = c("")) +
  tm_view(set.zoom.limits = c(11,17)) +
  tm_borders(alpha=0.5)

tmap_arrange(gdppc, LISAmap, asp=1, ncol=2)
```

## Hot Spot and Cold Spot Area Analysis

### Getis and Ord's G-Statistics

The analysis consists of three steps:

-   Deriving spatial weight matrix

-   Computing Gi statistics

-   Mapping Gi statistics

### Deriving distance-based weight matrix

We need to define a new set of neighbours. Whist the spatial autocorrelation considered units which shared borders, for Getis-Ord we are defining neighbours based on distance.

There are two type of distance-based proximity matrix, they are:

-   fixed distance weight matrix; and

-   adaptive distance weight matrix.

#### Deriving the centroid

To get our longitude values we map the *st_centroid()* function over the geometry column of us.bound and access the longitude value through double bracket notation \[\[\]\] and 1. This allows us to get only the longitude, which is the first value in each centroid.

```{r}
longitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[1]])

```

We do the same for latitude with one key difference. We access the second value per each centroid with \[\[2\]\].

```{r}
latitude <- map_dbl(hunan$geometry, ~st_centroid(.x)[[2]])
```

We use cbind to put longitude and latitude into the same object.

```{r}
coords <- cbind(longitude, latitude)
```

#### Determine the cut-off distance

We need to determine the upper limit for distance band by using the steps below:

-   Return a matrix with the indices of points belonging to the set of the k nearest neighbours of each other by using [*knearneigh()*](https://r-spatial.github.io/spdep/reference/knearneigh.html) of **spdep**.

-   Convert the knn object returned by *knearneigh()* into a neighbours list of class nb with a list of integer vectors containing neighbour region number ids by using [*knn2nb()*](https://r-spatial.github.io/spdep/reference/knn2nb.html).

-   Return the length of neighbour relationship edges by using [*nbdists()*](https://r-spatial.github.io/spdep/reference/nbdists.html) of **spdep**. The function returns in the units of the coordinates if the coordinates are projected, in km otherwise.

-   Remove the list structure of the returned object by using [**unlist()**](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/unlist).

```{r}
#coords <- coordinates(hunan)
k1 <- knn2nb(knearneigh(coords))
k1dists <- unlist(nbdists(k1, coords, longlat = TRUE))
summary(k1dists)
```

#### Computing fixed distance weight matrix

We will compute the distance weight matrix by using [*dnearneigh()*](https://r-spatial.github.io/spdep/reference/dnearneigh.html) as shown in the code chunk below.

```{r}
wm_d62 <- dnearneigh(coords, 0, 62, longlat = TRUE)
wm_d62

```

*nb2listw()* is used to convert the nb object into spatial weights object.

```{r}
wm62_lw <- nb2listw(wm_d62, style = 'B')
summary(wm62_lw)
```

### Computing adaptive distance weight matrix

It is possible to control the numbers of neighbours directly using k-nearest neighbours, either accepting asymmetric neighbours or imposing symmetry as shown in the code chunk below.

```{r}
knn <- knn2nb(knearneigh(coords, k=8))
knn
```

*nb2listw()* is used to convert the nb object into spatial weights object.

```{r}
knn_lw <- nb2listw(knn, style = 'B')
summary(knn_lw)
```

## Computing Gi statistics

### Gi statistics using fixed distance

```{r}
fips <- order(hunan$County)
gi.fixed <- localG(hunan$GDPPC, wm62_lw)
gi.fixed
```

We will join the Gi values to their corresponding hunan sf data frame by using the code chunk below.

```{r}
hunan.gi <- cbind(hunan, as.matrix(gi.fixed)) %>%
  rename(gstat_fixed = as.matrix.gi.fixed.)
```

The code chunk above performs three tasks. First, it convert the output vector (i.e. *gi.fixed*) into r matrix object by using *as.matrix()*. Next, *cbind()* is used to join hunan\@data and *gi.fixed* matrix to produce a new SpatialPolygonDataFrame called *hunan.gi*. Lastly, the field name of the gi values is renamed to *gstat_fixed* by using *rename()*.

### Mapping Gi values with fixed distance weights

The code chunk below shows the functions used to map the Gi values derived using fixed distance weight matrix.

```{r}
gdppc <- qtm(hunan, "GDPPC")

Gimap <-tm_shape(hunan.gi) +
  tm_fill(col = "gstat_fixed", 
          style = "pretty",
          palette="-RdBu",
          title = "local Gi") +
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, Gimap, asp=1, ncol=2)
```

### Gi statistics using adaptive distance

The code chunk below are used to compute the Gi values for GDPPC2012 by using an adaptive distance weight matrix (i.e *knb_lw*).

```{r}
fips <- order(hunan$County)
gi.adaptive <- localG(hunan$GDPPC, knn_lw)
hunan.gi <- cbind(hunan, as.matrix(gi.adaptive)) %>%
  rename(gstat_adaptive = as.matrix.gi.adaptive.)
```

### Mapping Gi values with adaptive distance weights

The code chunk below shows the functions used to map the Gi values derived using fixed distance weight matrix.

```{r}
gdppc<- qtm(hunan, "GDPPC")

Gimap <- tm_shape(hunan.gi) + 
  tm_fill(col = "gstat_adaptive", 
          style = "pretty", 
          palette="-RdBu", 
          title = "local Gi") + 
  tm_borders(alpha = 0.5)

tmap_arrange(gdppc, 
             Gimap, 
             asp=1, 
             ncol=2)
```
