---
title: "Hands-on Exercise 1: Geospatial Data Wrangling with R"
editor: visual
---

## Overview

In this hands-on exercise, I learn how to import and wrangle geospatial data using appropriate R packages.

## Getting Started

The code chunk below installs and loads [sf](https://r-spatial.github.io/sf/) and [tidyverse](https://www.tidyverse.org/packages/) packages into R environment.

```{r}
pacman::p_load(sf, tidyverse)
```

## Importing Geospatial Data

### Importing polygon feature data in Shapefile format

```{r}
mpsz <- st_read(dsn = "data/geospatial", layer = "MP14_SUBZONE_WEB_PL")
```

### Importing Polyline feature data in Shapefile Format

```{r}
cyclingpath <- st_read(dsn = "data/geospatial", layer = "CyclingPath")
```

### Importing GIS feature data in KML format

```{r}
preschool = st_read("data/geospatial/pre-schools-location-kml.kml")
```

## Checking Content of Simple Feature Data Frame

### Working with st_geometry()

```{r}
st_geometry(mpsz)
```

### Working with glimpse()

```{r}
glimpse(mpsz)
```

### Working with head()

```{r}
head(mpsz, n=5)
```

## Plotting Geospatial Data

### Plotting in R Graphic

```{r}
plot(mpsz)
```

### Plotting by Geometry

```{r}
plot(st_geometry(mpsz))
```

### Plotting by Specific Attribute

```{r}
plot(mpsz["PLN_AREA_N"])
```

## Working with Projection

### Assigning EPSG code to a simple data frame

```{r}
st_crs(mpsz)

mpsz3414 <- st_set_crs(mpsz, 3414) 

st_crs(mpsz3414)

```

### Transforming projection of preschool from wgs84 to svy21

```{r}
preschool3414 <- st_transform (preschool, crs = 3414)

preschool3414
```

## Importing and Converting an Aspatial Data

### Importing Aspatial Data

```{r}

listings <- read_csv("data/aspatial/listings.csv")

list(listings)
```

### Creating a simple feature data frame from aspatial data frame

```{r}
listings_sf <- st_as_sf(listings, coords = c("longitude", "latitude"), crs=4326) %>%
  st_transform(crs = 3414)

glimpse(listings_sf)
```

## Geoprocessing with sf package

### Buffering

```{r}
buffer_cycling <- st_buffer(cyclingpath, dist=5, nQuadSegs = 30)

buffer_cycling$AREA <- st_area(buffer_cycling)

sum(buffer_cycling$AREA)
```

### Point in polygon count

```{r}
mpsz3414$`PreSch Count`<- lengths(st_intersects(mpsz3414, preschool3414)) 

summary(mpsz3414$`PreSch Count`)

top_n(mpsz3414, 1, `PreSch Count`)

mpsz3414$Area <- mpsz3414 %>%
  st_area()

mpsz3414 <- mpsz3414 %>%
  mutate(`PreSch Density` = `PreSch Count`/Area * 1000000)

```

## Exploratory Data Analysis (EDA)

### Plotting in Histogram

```{r}

hist(mpsz3414$`PreSch Density`)

ggplot(data=mpsz3414, 
       aes(x= as.numeric(`PreSch Density`)))+
  geom_histogram(bins=20, 
                 color="black", 
                 fill="light blue") +
  labs(title = "Are pre-school evenly distributed in Singapore?",
       subtitle= "There are many planning sub-zones with a single pre-school, on the other hand, \nthere are two planning sub-zones with at least 20 pre-schools",
      x = "Pre-school density (per km sq)",
      y = "Frequency")
```

### Plotting in Scatterplot

```{r}

mpsz3414

ggplot(mpsz3414, aes(x = as.numeric('PreSch Density'), y = as.numeric('PreSch Count'))) + 
  geom_point(size=2) +
  
  labs(title = "Is there a relationship between Pre-School Density and Pre-School Count?",
       x = "Pre-school density (per km sq)",
       y = "PreSch Count")
```
